---
title: "Basics"
format:
  html:
    self-contained: true
editor: visual
---

This document covers the basic concepts in R for the workshop *Projects in R: Creating Reproducible Analyses*. The workshop is tailored towards early-stage researchers in psychology, cognitive science and related disciplines. It assumes beginner to intermediate expertise with R.

# Preparation and Terminology

First, we install and load required packages to introduce basic concepts in R. *Note* that we will use the conventional (non project-based) installation here.

Libraries can be loaded using either `library` or `require`. These function work almost identical. The main differences is that `require` returns `FALSE` when the package is not installed whereas `library` will raise an error.

```{r}
# Install is only required if dplyr is not yet installed
# install.packages("renv")
# Load the package
library(renv)
```

When packages are attached R loads all functions into the namespace (e.g., when renv is loaded its functions activate, autoload etc. become directly available). When the name of a function has already been assigned, it will be masked. This means that the function from the last loaded package are directly callable, masked functions from the previously loaded packages have to be called from the package (via `::`).

```{r}
# Calling function explicitly from a package
base::getwd()
```

## Getting help

R has a great and clearly structured documentation. - Description - Usage - Arguments - Details - Value(s) - Examples Help for objects or functions can be called via the prefix `?`.

```{r echo=TRUE}
?library
```

# Presentation contents

# Basics of Working with R

## R

-   Programming language
-   Primarily used for statistical and graphics computing
-   Can be used through different applications
    -   Terminal/Console
    -   Integrated development environments (IDEs, e.g., VS Code)
    -   Interactive notebooks (Jupyter)

## Rstudio I

-   IDE specialized on R (and Python)
-   Collection of tools that enhance work with R
    -   Console
    -   Editor with syntax highlighting and code completion
    -   Graphing tools
    -   Git integration

## Rstudio II

-   Consists of multiple panes/panels
    -   Source/Script
    -   Console and Terminal
    -   2 panes with further functionality
        -   Environment
        -   Files
        -   Git
        -   Help
        -   Packages
        -   Plots
-   Layout can be customized via `Tools > Global Options > Pane Layout`

## Console and Terminal

-   Console refers to the R console
    -   Equivalent to opening R in a terminal window
    -   Allows direct interaction with R
        -   E.g., Complete mathematical operations: `1 + 2 - 3 * 4 / 5`
        -   Has history but all interactions are lost upon closing RStudio
    -   Terminal lets you interact with your operating systems shell
        -   Mac/Linux: typically `bash` or `zsh`
        -   Windows: `PowerShell`
        -   Can be used for manual git integration

## Source/Scripts

-   Scripts are edited here
    -   They are text files and can be saved
-   Has incorporated highlighting and formatting tools
    -   e.g., `Cmd/Ctrl + Shift + A` formats code in R conventions
-   The primary way we write code
-   R Studio Code lets you seamlessly run this code
    -   Graphical User Interface: `Run Button`
    -   Keyboard Shortcuts: `Cmd/Ctrl + Return`
    -   Selected code is send to the console and evaluated

## Further Panes I

-   Environment
    -   Overview of objects on the working environment
    -   Useful to inspect data, models etc.
    -   GUI tool for data import
-   Files
    -   File manager
-   Help
    -   Documentation and help packages
    -   Documentation will be displayed here if help is called from r (`?function_name`)

## Further Panes II

-   Packages
    -   Overview of installed and loaded packages
    -   Can be used to load and update packages via GUI
-   Plots
    -   Graphs created in R are displayed here
    -   GUI options for export etc.
-   Git
    -   Version control
    -   A GUI tool with most common git commands

## Hidden files and folders

-   Most of the tools we cover use hidden files or folders
-   Works similar across file systems (`Mac: Finder`, `Win: Explorer`)
    -   Per default elements with a leading `.` will not be displayed
    -   Can be shown via shortcut (`Cmd + Shift + .`, `Alt + V` then `H + H`)
-   Commonly used by applications to save settings
    -   Called dotfiles
    -   E.g., Rstudio settings on Mac are saved under `~/.config/rstudio/rstudio-prefs.json`

# RStudio walkthrough

# R

## Working with R

> To understand computations in R, two slogans are helpful:
>
> -   Everything that exists is an object.
> -   Everything that happens is a function call. [John Chambers (creator of the S programming language)](https://riegersven.github.io/data-doc-workshop-2023/intro-r-rstudio.html)

## Data Structures

-   Vectors
-   Lists
-   Data Frames

## Vectors

-   One-dimensional array with one data type
-   Created wit `c()` (combine) function

```{r}
#| echo: true
c(1, 2, 3, 4)
```

-   Elements can be assessed via their index
    -   R starts indices with 1

```{r}
#| echo: true
c(1:4)[3]
```

## Named Vectors

-   Make elements accessible by name
-   Getting the value(s) requires extra steps

```{r}
#| echo: true
one_to_three <- c("one" = 1, "two" = 2, "three" = 3)
one_to_three
```

```{r}
#| echo: true
# This doesn't work
# one_to_three["two"] + 1
# With name
unlist(one_to_three["two"]) + 1
# Without name
unname(one_to_three["two"]) + 1
```

## Lists

-   List
    -   (Named) collection of objects
    -   Can contain multiple data types
    -   Objects can be accessed via their position or name
    -   Resembles dictionaries from other languages (e.g., python)
    -   Commonly used for results of analyses

```{r}
#| echo: true
list(`1` = "one", "range" = c(2:4), "matrix" = matrix(1:4, nrow = 2))
```

## Data Frames

-   Data Frame
    -   Named table
    -   Multiple data types
        -   Consistent within columns (e.g., one non-numeric value $\rightarrow$ character column)
    -   Columns can be accessed via their name (`$`, `[]` or `.`)
    -   Rows and columns can be accessed via their index

```{r}
#| echo: true
numbers <- data.frame("number" = c(1:4), "letter" = c("a", "b", "c", "d"))
numbers$number
numbers[, "letter"]
numbers[4, 2]
```

## Functions I

-   A function is a bundled unit of code that performs a specific task
-   Functions generally
    -   Have a name
    -   Take arguments
    -   Return (a) value(s)
-   In plain terms: a function uses a predefined procedure to create (a) value(s) based on (an) argument(s).

## Functions II

-   Advantages of functions
-   Functions R are a great way to improve your R workflows
    -   Repeat tasks without risk of copy and paste errors
    -   When the task needs adjustments, changes only need to be made in one place
    -   Objects and variables defined within a function are kept within the function scope unless specified differently
    -   Improve readability of your code

## Functions III

-   Are created with the `function` keyword
-   Can have no arguments
-   Arguments can have default values
-   It is considered good practice to always include a return value
    -   If none is required the function can still return `TRUE` or `NULL`

```{r}
#| echo: true
say_hello <- function() {
  return("Hello, I am a function!")
}
say_hello()

say_hello_to <- function(who = "world"){
  hello_to <- paste("Hello, ", who, "!", sep = "")
  return(hello_to)
} 
say_hello_to()
say_hello_to("workshop attendees")
```

## Additional Information

In this example we create a function that adds one to a given number. It's only argument is the input number and it return value is the output number (= input number + 1).

```{r}
add_one <- function(input_number) {
  output_number = input_number + 1
  return(output_number)
}

# We can pass the argument with or without the keyword
add_one(1)
add_one(input_number = 2)
```

An advantage of functions is that the series of operations can be repeated easily. Here is an example for all numbers between 1 and 10 written in a `for loop`. We first create an empty `vector` using the combine function (`c()`). For each number between one and ten, we then add the value from add_one to our vector. In R elements are added to a vector by combining the vector and the new element. We further save the times befor and after running the loop to compare it to the method below.

```{r}
start_time_loop <- Sys.time()
two_to_eleven <- c()
for (i in 1:10) {
  two_to_eleven <- c(two_to_eleven, add_one(i))
}
stop_time_loop <- Sys.time()
two_to_eleven
```

However, while `for` or `while` loops are intuitive, they are typically not the most suitable way to perform functions on multiple values. Because of memory allocation advantages it is generally faster and more robust to apply/map the function to all values. This is often referred to as vectorization. R has a variety of so called wrappers implemented for the `apply` function. Here we use `sapply` to execute the function for each number between one and ten. *Note* that we delete two_to_eleven with the `rm` function first.

```{r}
rm(two_to_eleven)
start_time_apply <- Sys.time()
two_to_eleven <- sapply(1:10, add_one)
stop_time_apply <- Sys.time()
two_to_eleven
```

Even in this very small example we can already see a sizable performance difference between the loop and apply approach.

```{r}
loop_time <- stop_time_loop - start_time_loop
apply_time <- stop_time_apply - start_time_apply
loop_time
apply_time
```

When we define a function, we can further define default values for any of the arguments it has by simple adding = and the default value to the argument. Unless a value for this argument is provided when this function is called, it will resort to the default value for the argument.

```{r}
rounded_mean_sd <- function(vec, digits = 2) {
  mean_r = round(mean(vec), digits)
  sd_r = round(sd(vec), digits)
  rounded_string = paste(mean_r, " (", sd_r, ")", sep = "")
  return(rounded_string)
}
# Draw a radom sample from N(2/3, 1/6)
test_data <- rnorm(100, 2/3, 1/6)
rounded_mean_sd(test_data)
rounded_mean_sd(test_data, digits = 3)
rounded_mean_sd(test_data, digits = 6)
```

# Function excercise

Create a function that returns a named vector or list with the mean, sd, min and max of vector.
