---
title: "Computational Reproducibility"
format:
  html:
    self-contained: true
editor: visual
---

## Random Number Generation

-   Some models and functions involve random processes
    -   Results change when repeated
    -   E.g., splitting data into n random subsets to cross-validate an exploratory factor analysis
-   Example: drawing 10 values from a Z-Distribution

```{r}
#| echo: true
rnorm(10)
rnorm(10)
```

## Setting seeds

-   Sets a "fixed" seed that makes random processes reproducible

```{r}
#| echo: true
set.seed(1)
rnorm(10)
set.seed(1)
rnorm(10)
```

-   Use random seeds when reporting such results
-   Make sure that the results do not vastly change when seed is changed
    -   E.g., test model/function several time without or with different random seeds
-   If there are substantial fluctuations, repeat process several times
    -   Aggregate results or report fluctuation

# Structuring Projects

## Paths

-   Absolute paths make it impossible to run script on other machines without changing them (e.g., "/Users/this/path/wont/work/for_you.csv")
-   Key advantage of R projects: Console and terminal automatically start in project folder
    -   Relative paths can be used (e.g., "./data/my_example_file.csv")
-   `here` package can be used to overcome differences in OS

{.fragement}

```{r}
#| echo: true
library(here)

# Create compatible path for current OS
file_path <- here("data", "raw", "example_logfile.csv")
file_path
# Could be used to load a file
# log_file <- read.csv(file_path, stringsAsFactors = FALSE)
```

## Folder Structure in Projects I

-   Hard to find relevant files if all files and scripts are at the top level of your folder
-   Projects should follow clear structure
-   Example
    -   Project
        -   Data
            -   Raw
            -   Processed
        -   Scripts
        -   Output
            -   Graphs
            -   Tables

## Folder Structure in Projects II

-   The structure depends on your preferences and the project
-   Recommendation: Keep to level as clean as possible (e.g., only folders, README.md etc.)
-   [Resource](https://www.youtube.com/watch?v=GeN-qqNLLsM) that explains three ways to create the folder structure
    -   In your file explorer
    -   Using R
    -   From the terminal

{.callout-note} Avoid nesting projects within a projects.

## Organizing Scripts

-   All libraries should be loaded at the start of the script
    -   Loading packages just before they are used is error prone when changes are made
-   There is not agreed upon convention for their order
    -   In order of their first use
    -   Alphabetical
    -   With masking in mind (e.g., load packages whose functions should not be masked last)

## Splitting Scripts and Sourcing

-   Scripts should be split if they get extensive and include multiple key steps
    -   E.g., Functions, data import and pre-processing, analysis
-   `source` if they create multiple objects necessary for subsequent steps
    -   Masking and similar issues may occur
-   If they only create a data set (e.g., during pre-processing) saving the result of that script as a file is preferable
-   Objects can also be saved as R data
    -   For large computations that can't be saved in standard file formats

# Improving Code Readability

## Piping and Nesting

-   Piping: "Chaining" operations through a pipe operator
-   The value of the left-hand side is used as the first argument of the right-hand side
    -   Placeholder can be used to use value in as different argument in rhs
-   Two implementations
    -   `%>%` (`dplyr`/`tidyverse`)
    -   `|>` (Base R [version 4.1.0](https://cran.r-project.org/doc/manuals/r-devel/NEWS.html) or later)
-   Can greatly improve readability of nested function call
    -   Ideal for data manipulation

## Nesting - Example

```{r}
library(dplyr)
```

-   Can you quickly tell what this code does?

{.fragment}

```{r}
#| echo: true
summarise(
  group_by(filter(iris, Species != "setosa"), Species),
  sepal_length = mean(Sepal.Length),
  petal_width = mean(Petal.Width)
)
```

## Piping - Example

-   How about now?

```{r echo=TRUE}
#| echo: true
iris %>%
  filter(Species != "setosa") %>%
  group_by(Species) %>%
  summarise(sepal_length = mean(Sepal.Length),
            petal_width = mean(Petal.Width))
```

-   Pipe-related resources
    -   [dplyr](https://dplyr.tidyverse.org/): powerfull data manipluation with pipes
    -   [rstatix](https://rpkgs.datanovia.com/rstatix/): common statistical models with focus on piping

## Commenting Code

-   Comments are lines in scripts that are not evaluated
-   IDEs (e.g., RStudio) will accentuate comments
-   R + RStudio
    -   In R comments start with `#`
    -   Keyboard shortcut `Cmd/Ctrl + Shift + C`
-   Comments can be used between or at the end of lines

```{r}
# Show that statement if not printed when commented out
# print("This is an example for a comment")
print("That comment didn't stop me!") # Code not 'disabled'
```

## Importance of Comments

-   Readable and comprehensible code is essential for a reproducible workflow
-   Comments are essential to efficiently resume your projects
-   Enable others to work with your code more easily
-   Commenting your code clearly takes practice and effort

## Ideas to Effectively Comment

-   There are no clear guidelines for commenting
-   Be concise and precise
-   Point out specifics that are not easily detectable
-   State what a code snippet does

```{r}
#| echo: true
# Get the absolute path of this file via Rstudioapi
# Only run if rstudioapi is installed
# library(rstudioapi)
# if (require(rstudioapi)){
#   dirname(rstudioapi::getSourceEditorContext()$path)
# }
```

-   Don't state obvious things (the documentation handled that)
-   Point out dependencies
-   Find a commenting style that works for you

## Naming

-   Often overlooked when improving coding
-   Less need to limit length of variable names with help of development tools (i.e., code completion)
-   Variable and object naming makes code more readable
    -   Can work like comments
-   Common conventions
    -   camelCase
    -   Use_of_underscores
    -   In R: `.` is often used instead of `_`
        -   e.g., `read.csv`
        -   Clashes with naming conventions in other languages, particularly object-oriented languages (e.g. python)
-   Consistency is important
    -   Functions with similar functionality should follow the same convention
    -   Different conventions can be used for different objects (e.g., [Julia blue coding style]())

## Outlook

-   Project-based workflows are applicable to most languages
-   Python
    -   [Spyder](https://www.spyder-ide.org/) is it's equivalent to RStudio
    -   [venv](https://docs.python.org/3/library/venv.html) works very similar to `renv`
        -   Installing are still made with `pip`, but the libraries are saved in the virtual environment
    -   [pyenv](https://github.com/pyenv/pyenv) can further change Python versions easily
-   Julia
    -   [VSCode](https://code.visualstudio.com/) is the officially supported IDE
    -   Has a built in project mode (`julia -project`)
    -   [Julia-up](https://github.com/JuliaLang/juliaup) enables easy switching between Julia versions
