---
title: "Technical Reproducibility"
format:
  html:
    self-contained: true
editor: visual
---

## Package Management - Relevance I

-   Most analyses you will conduct will require the installation of additional packages
    -   Developed decentralized by different groups
    -   Often require functions from one another (dependencies)
    -   Updated over time in vastly differing intervals
-   Potential issues
    -   Required packages can be incompatible
        -   E.g., two packages require different versions of a dependency
    -   Functions were changed $\rightarrow$ old scripts don't run after update
    -   Default values are changed $\rightarrow$ results change
    -   R's functionality gets updated $\rightarrow$ packages may not work at all

## Package Management - Relevance II

-   Reproducibility
    -   Sharing data and scripts is at the core of open science efforts
    -   Hardly achievable for analysis when script can't be run
        -   On other machines
        -   At later time points
-   Save unnecessary effort when
    -   Reusing parts of old scripts for new projects
    -   Share scripts with colleagues

## Projects

-   A solution to most of the aforementioned issues
-   Idea: Separate package libraries for each project
    -   Only install required packages for a project
    -   Breakage in a project does not affect other projects
    -   Slimmer libraries $\rightarrow$ package conflicts less likely
-   A common approach in programming-related work
    -   Most IDEs include useful tools for project-based workflows

## What is renv?

-   R package that helps to manage package versions and dependencies in projects
-   Aims
    -   Enable robust, local projects
    -   Enhance reproducibility
-   Key functionalities
    -   Automatically documents installed package versions
    -   Provides a means to revert project-breaking changes
    -   Scans project for required packages and advises installation of missing packages
    -   Tools for package management (e.g., convenient way to update all packages)

## How does `renv` Work?

-   renv creates key files and folder when initiated
    -   renv folder
        -   Sub folders for libraries and settings
        -   `activate.R` script, which makes sure that the local project library is used for the R session when its executed
    -   `.Rprofile`
        -   File that can be used to configure r functionality as/before it starts
        -   Here sources("renv/activate.R")
    -   Version of installed packages is saved in `renv.lock`

## Project-based Workflow with `renv`

1.  Set up renv

```{r}
#| echo: true
#| eval: false
renv::init()
```

2.  Work on project (install packages etc)

```{r}
#| echo: true
#| eval: false
renv::install("package I need")
```

3.  Save the packages (incl. version used)

```{r}
#| echo: true
#| eval: false
renv::snapshot()
```

4.  Change/update packages when necessary

-   Successful: repeat 3
-   Broken project: renv::revert to restore last snapshot

```{r}
#| echo: true
#| eval: false
renv::revert()
```

## Further Useful Tools

-   Check the status of the current environment
    -   Changes that are not yet recorded in `renv.lock`
    -   Packages that are required but not installed

```{r}
#| echo: true
#| eval: false
renv::status()
```

-   Detect packages that are required

```{r}
#| echo: true
#| eval: false
renv::dependencies()
```

-   Copy required packages from local R installation rather than downloading them new

```{r}
#| echo: true
#| eval: false
renv::hydrate()
```

# `renv` Demo and Exercise

## `renv` Exercise

::: callout-warning
Don't use your home directory or an existing project folder for this exercise.
:::

-   Initialize a new r environment in a folder new folder
-   Create an r script in the folder
    -   Add code to load two uninstalled libraries
    -   Don't execute that script
    -   Close the script
-   Check the dependencies of the renv
-   Install libraries and snapshot the environment
-   Install an unnecessary library and revert the install
